BlueTracer: \\ a Robust API Tracer for Evasive Malware

### 2018-07-09 background.tex
First Batch of fixes (Emilio's)

\chapter{Background}

@@

The main techniques which use the first approach are \textit{Import Address Table (IAT) Patching}, \textit{Export Address Table (EAT) patching} and \textit{Proxy DLL}. On the other hand, the most significant techniques which use the second approach are \textit{inline hooking} and \textit{debugger-based hooking} (\textit{Figure 1.1}).

\begin{figure}[h]
\centering
\hspace*{-3em}
\includegraphics[scale=0.8]{Figures/API-hooking-3.pdf}
\caption{\textit{API hooking techniques classification}}
\end{figure}

@@

In the header of every Portable Executable (PE) file there is an Import Address Table (IAT) for every dynamic-link library (DLL) that is included by the executable \cite{Berdajs:2010:EAU:1815744.1815746} (\textit{Figure 1.3}). This table is utilized to indicate the location of DLL-imported functions in virtual memory and is filled by the Windows loader with the actual function memory addresses after the executable is loaded in memory.

\begin{figure}[h]
\centering
\hspace*{-3em}
\includegraphics[scale=0.8]{Figures/IAT-3.pdf}
\caption{\textit{IAT in PE header}}
\end{figure}

@@

In \textit{inline hooking} the API to be hooked has its initial instructions (at least the first 5 bytes) overwritten with an unconditional jump to a replacement function. In order to ensure that the API's original functionality is not lost due to the modification of its entry point, a \textit{trampoline function} is created, consisting of a copy of the overwritten instructions and an unconditional jump back to the unaltered portion of the original function. As a result of this, the replacement function can invoke the original function by calling the trampoline, after performing all the desired analysis operations \cite{Berdajs:2010:EAU:1815744.1815746}.
\textit{Figure 4} illustrates a program's execution flow before and after the use of \textit{inline hooking}.

\begin{figure}[h]
\centering
\includegraphics[scale=0.8]{Figures/Inline.pdf}
\caption{\textit{(a) Ordinary API call execution flow \\
				  (b) API call with inline hooking}}
\end{figure}

>>>>

\chapter{Background} \label{background}

@@

The main techniques which use the first approach are \textit{Import Address Table (IAT) Patching}, \textit{Export Address Table (EAT) patching} and \textit{Proxy DLL}. On the other hand, the most significant techniques which use the second approach are \textit{inline hooking} and \textit{debugger-based hooking} (\textit{Figure~\ref{HookTech}}).

\begin{figure}[h]
\centering
\hspace*{-3em}
\includegraphics[scale=0.8]{Figures/API-hooking-3.pdf}
\caption{\textit{API hooking techniques classification}}
\label{HookTech}
\end{figure}

@@

In the header of every Portable Executable (PE) file there is an Import Address Table (IAT) for every dynamic-link library (DLL) that is included by the executable \cite{Berdajs:2010:EAU:1815744.1815746} (\textit{Figure~\ref{IAT}}). This table is utilized to indicate the location of DLL-imported functions in virtual memory and is filled by the Windows loader with the actual function memory addresses after the executable is loaded in memory.

\begin{figure}[h]
\centering
\hspace*{-3em}
\includegraphics[scale=0.8]{Figures/IAT-3.pdf}
\caption{\textit{IAT in PE header}}
\label{IAT}
\end{figure}

@@

In \textit{inline hooking} the API to be hooked has its initial instructions (at least the first 5 bytes) overwritten with an unconditional jump to a replacement function. In order to ensure that the API's original functionality is not lost due to the modification of its entry point, a \textit{trampoline function} is created, consisting of a copy of the overwritten instructions and an unconditional jump back to the unaltered portion of the original function. As a result of this, the replacement function can invoke the original function by calling the trampoline, after performing all the desired analysis operations \cite{Berdajs:2010:EAU:1815744.1815746}.
\textit{Figure~\ref{Inlined}} illustrates a program's execution flow before and after the use of \textit{inline hooking}.

\begin{figure}[h]
\centering
\includegraphics[scale=0.8]{Figures/Inline.pdf}
\caption{\textit{(a) Ordinary API call execution flow \\
				  (b) API call with inline hooking}}
\label{Inlined}
\end{figure}

### 2018-07-09 evaluation.tex
First Draft Corrected WOUT JoeBox
First Batch of fixes (Emilio's)

\chapter{Experimental Evaluation}

@@

In order to evaluate the run-time overhead introduced by BlueTracer we tested it with a number of benign applications exercising a large number of functions. In order to experiment with a wide range of different functionalities, we picked a set of well-known Windows applications performing a variety of jobs. We grouped these applications based on the task they carry out, as shown below:

\begin{itemize}
\item \textbf{Collection of system information}
	   \begin{itemize}
	   \item \texttt{Systeminfo.} It outputs a summary of OS-environment parameters. 
	   \item \texttt{System File Checker (SFC).} It scans for corruptions in system files.
	   \item \texttt{Check Disk}. It checks the disks' integrity.
	   \item \texttt{IPConfig}. It displays TCP/IP network configuration values.
	   \item \texttt{Netstat}. It can show active TCP connections as well as TCP and UDP ports on which the computer is listening.
	   \item \texttt{Driver Query}. It presents the list of installed drivers and their properties.
	   \item \texttt{Windows Assessment Tools (WinSAT).} It  measures a number of performance characteristics and reports them.
	   \item \texttt{Powercfg.} It enables to search for common energy-efficiency problems. 
	   \end{itemize}
\item \textbf{Compression programs}
		\begin{itemize}
		\item \texttt{7zip.}. Its default output extension is \texttt{.7z}.
		\item \texttt{IZArc.} Its default output extension is \texttt{.zip}.
		\item \texttt{WinRar.} Its default output extension is \texttt{.rar}.
		\end{itemize}
\item \textbf{Encryption Utilities}
		\begin{itemize}
		\item \texttt{Cipher.} It uses Encrypting File System (ESF), based on DESX. 
		\item \texttt{OpenSSL.} It was configured to employ 256-bit CBC AES.
		\item \texttt{Crypt.} It adopts RC2 block encryption.
		\end{itemize}
\item \textbf{Hashing Utilities}
		\begin{itemize}
		\item \texttt{File Checksum Integrity Verifier (FCIV).} It was set to use both MD5 and SHA1.
		\item \texttt{TurboSFV.} It was configured to adopt SHA3-224. 
		\end{itemize}
\end{itemize}

@@

\hline
\hline
   \textbf{Application} & \specialcell{ \textbf{Native} \\ \textbf{(s)}} & \specialcell{ \textbf{Empty Image} \\ \textbf{(s)}} & \specialcell{ \textbf{Empty Routine} \\ \textbf{(s)}} & \specialcell{ \textbf{Main Image} \\ \textbf{(s)}} & \specialcell{ \textbf{Complete} \\ \textbf{(s)}} \\
\hline
\texttt{Systeminfo}      & 5.894    & 9.686  & 12.869 & 15.694 & 26.398     \\
\texttt{SFC}          & 218.251 & 222.065 & 226.125 & 230.869 & 245.476       \\
\texttt{Check Disk}       & 101.570     & 171.94 & 181.429 & 454.649 & 600+      \\
\texttt{IPConfig}       & 0.105     & 4.257 & 8.421 & 9.599 & 13.181      \\
\texttt{Netstat} & 0.085 & 2.563 & 4.630 & 6.902 & 21.129       \\
\texttt{DriverQuery} & 1.168 & 6.663 & 9.786 & 34.074 & 68.668      \\
\texttt{WinSat}       & 163.175     & 212.496 & 221.012 & 273.641 & 600+      \\
\texttt{PowerCFG} & 61.333      & 73.325 & 80.018 & 132.466 & 273.882       \\
\texttt{7zip}       & 175.526     & 212.000 & 225.091 & 227.985 & 233.792      \\
\texttt{IZArc}      & 30.920     & 52.878 & 56.631 & 61.066 & 65.770      \\
\texttt{WinRar} & 155.674 & 192.638 & 198.804 & 208.606 & 215.455      \\
\texttt{Cipher} & 0.028 & 1.817 & 3.787 & 4.435 & 5.064       \\
\texttt{OpenSSL} & 24.769 & 33.183 & 34.308 & 47.006 & 1524.763      \\
\texttt{Crypt} & 65.392 & 140.048 & 142.979 & 572.353 & 600+      \\
\texttt{FCIV} & 10.807 & 25.113 & 28.144 & 56.329 & 87.948       \\
\texttt{TurboSFV} & 34.412 & 57.705 & 62.639 & 66.439 & 74.657
\\
\hline
\end{tabularx}
\vspace{0.2cm}
\caption{Execution times of instrumented benign applications}

@@

\hline
\hline
\cline{1-2}
   \textbf{Application} & \specialcell{ \textbf{Native} \\ \textbf{(s)}} & \specialcell{ \textbf{Complete} \\ \textbf{(s)}} & \textbf{Events} & \specialcell{ \textbf{Overhead per Event} \\  \textbf{($10^-5$ s)} }  \\
\hline
\texttt{Systeminfo}      & 5.894    & 26.398  & 387 476 & $5.291$    \\
\texttt{SFC}          & 218.251 & 245.476 & 333 781 & $8.156$ \\
\texttt{IPConfig}       & 0.105 & 13.181 & 94 313 & $13.864$      \\
\texttt{Netstat} & 0.085 & 21.129 & 542 649 & $3.878$       \\
\texttt{DriverQuery} & 1.168 & 68.668 & 1 735 859 & $3.860$            \\
\texttt{PowerCFG} & 61.333 & 273.882 & 4 497 247 & $4.726$       \\
\texttt{7zip}       & 175.526 & 233.792 & 146 578 & $39.751$      \\
\texttt{IZArc}      & 30.920 & 65.770 & 109 142 & $31.931$      \\
\texttt{WinRar} & 155.674 & 215.455 & 84 329 & $70.890$      \\
\texttt{Cipher} & 0.028 & 4.435 & 24 141 & $18.255$           \\
\texttt{OpenSSL} & 24.769 & 1524.763 & 57 410 809 & $2.613$                \\
\texttt{FCIV} & 10.807 & 87.948 & 725 133 &  $10.638$               \\
\texttt{TurboSFV} & 34.412 & 74.657 & 183 455 & $21.937$ 
\\
\hline
\end{tabularx}
\vspace{0.2cm}
\caption{Run-time overhead for benign applications}

@@

Lastly, anti-analysis checks are aimed at detecting the existence processes belonging to popular analysis tools. One of such tools is Process Monitor (\texttt{procmon.exe}), which is an advanced Windows monitoring tool that shows real-time file system, registry and process/thread activity \cite{procmon}. Al-Khaser looks for the processes of many analysis tools, including the process associated to Process Monitor. Again, BlueTracer was also able to successfully log anti-analysis events, even the ones directed at uncovering the presence of Process Monitor (\textit{Listing 4.11}). From the listing below, it can be seen that the course of action adopted by Al-Khaser is quite straightforward. It first takes a snapshot of all the processes in the system with \texttt{CreateToolhelp32Snapshot}. Then, for each process in the snapshot, it retrieves the corresponding name, through the use of \texttt{Process32FirstW} at the beginning and \texttt{Process32NextW} afterwards, and compares it with \texttt{procmon.exe} via \texttt{StrCmpIW}.

>>>>

\chapter{Experimental Evaluation} \label{results}

@@

In order to evaluate the run-time overhead introduced by BlueTracer we tested it with a number of benign applications exercising a large number of functions. We had to perform this kind of test on deterministic batch programs to obtain accurate results rather than on real malware samples. This because malware instances have unpredictable behaviour, thus rendering the execution times obtained when running them meaningless.
Having in mind the goal of experimenting with a wide range of different functionalities, we picked a set of well-known Windows applications performing a variety of jobs. We grouped these applications based on the task they carry out, as shown below:

\begin{itemize}
\item \textbf{Collection of system information}
	   \begin{itemize}
	   \item \texttt{Systeminfo.} It outputs a summary of OS-environment parameters. 
	   \item \texttt{System File Checker (SFC).} It scans for corruptions in system files.
	   \item \texttt{Check Disk}. It checks the disks' integrity.
	   \item \texttt{IPConfig}. It displays TCP/IP network configuration values.
	   \item \texttt{Netstat}. It can show active TCP connections as well as TCP and UDP ports on which the computer is listening.
	   \item \texttt{Driver Query}. It presents the list of installed drivers and their properties.
	   \item \texttt{Windows Assessment Tools (WinSAT).} It  measures a number of performance characteristics and reports them.
	   \item \texttt{Powercfg.} It enables to search for common energy-efficiency problems. 
	   \end{itemize}
\item \textbf{Compression programs}
		\begin{itemize}
		\item \texttt{7zip.} Its default output extension is \texttt{.7z}.
		\item \texttt{IZArc.} Its default output extension is \texttt{.zip}.
		\item \texttt{WinRar.} Its default output extension is \texttt{.rar}.
		\end{itemize}
\item \textbf{Encryption Utilities}
		\begin{itemize}
		\item \texttt{Cipher.} It uses Encrypting File System (ESF), based on DESX. 
		\item \texttt{OpenSSL.} It was configured to employ 256-bit CBC AES.
		\item \texttt{Crypt.} It adopts RC2 block encryption.
		\end{itemize}
\item \textbf{Hashing Utilities}
		\begin{itemize}
		\item \texttt{File Checksum Integrity Verifier (FCIV).} It was set to use both MD5 and SHA1.
		\item \texttt{TurboSFV.} It was configured to adopt SHA3-224. 
		\end{itemize}
\end{itemize}

@@

\hline
\hline
   \textbf{Application} & \specialcell{ \textbf{Native} \\ \textbf{(s)}} & \specialcell{ \textbf{Empty Image} \\ \textbf{(s)}} & \specialcell{ \textbf{Empty Routine} \\ \textbf{(s)}} & \specialcell{ \textbf{Main Image} \\ \textbf{(s)}} & \specialcell{ \textbf{Complete} \\ \textbf{(s)}} \\
\hline
\texttt{Systeminfo}      & 5.894    & 9.686 (1.64x)  & 12.869 (2.18x) & 15.694 (2.66x) & 26.398 (4.48x)     \\
\texttt{SFC}          & 218.251 & 222.065 (1.02x) & 226.125 (1.04x) & 230.869 (1.06x) & 245.476 (1.12x)       \\
\texttt{Check Disk}       & 101.570     & 171.94 (1.69x) & 181.429 (1.79x) & 454.649 (4.48x) & 600+       \\
\texttt{IPConfig}       & 0.105     & 4.257 (40.54x) & 8.421 (80.20x) & 9.599 (91.42x) & 13.181 (125.53x)      \\
\texttt{Netstat} & 0.085 & 2.563 (30.15x) & 4.630 (54.47x) & 6.902 (81.20x) & 21.129 (248.58x)       \\
\texttt{DriverQuery} & 1.168 & 6.663 (5.70x) & 9.786 (8.38x) & 34.074 (29.17x) & 68.668 (58.79x)     \\
\texttt{WinSat}       & 163.175     & 212.496 (1.30x) & 221.012 (1.35x) & 273.641 (1.68x) & 600+      \\
\texttt{PowerCFG} & 61.333      & 73.325 (1.20x) & 80.018 (1.30x) & 132.466 (2.16x) & 273.882 (4.47x)       \\
\texttt{7zip}       & 175.526     & 212.000 (1.21x) & 225.091 (1.28x) & 227.985 (1.30x) & 233.792 (1.33x)      \\
\texttt{IZArc}      & 30.920     & 52.878 (1.71x) & 56.631 (1.83x) & 61.066 (1.97x) & 65.770 (2.13x)      \\
\texttt{WinRar} & 155.674 & 192.638 (1.24x) & 198.804 (1.28x) & 208.606 (1.34x) & 215.455 (1.38x)      \\
\texttt{Cipher} & 0.028 & 1.817 (64.89x) & 3.787 (135.25x) & 4.435 (158.39x) & 5.064 (180.86x)       \\
\texttt{OpenSSL} & 24.769 & 33.183 (1.34x) & 34.308 (1.39x) & 47.006 (1.90x) & 1524.763 (61.56x)      \\
\texttt{Crypt} & 65.392 & 140.048 (2.14x) & 142.979 (2.19x) & 572.353 (8.75x) & 600+      \\
\texttt{FCIV} & 10.807 & 25.113 (2.32x) & 28.144 (2.60x) & 56.329 (5.21x) & 87.948 (8.14x)       \\
\texttt{TurboSFV} & 34.412 & 57.705 (1.68x) & 62.639 (1.82x) & 66.439 (1.93x) & 74.657 (2.17x)
\\
\hline
\end{tabularx}
\vspace{0.2cm}
\caption{Execution times of instrumented benign applications}

@@

\hline
\hline
   \textbf{Application} & \specialcell{ \textbf{Native} \\ \textbf{(s)}} & \specialcell{ \textbf{Complete} \\ \textbf{(s)}} & \textbf{Events} & \specialcell{ \textbf{Overhead per Event} \\  \textbf{($10^-5$ s)} }  \\
\hline
\texttt{Systeminfo}      & 5.894    & 26.398  & 387 476 & $5.291$    \\
\texttt{SFC}          & 218.251 & 245.476 & 333 781 & $8.156$ \\
\texttt{IPConfig}       & 0.105 & 13.181 & 94 313 & $13.864$      \\
\texttt{Netstat} & 0.085 & 21.129 & 542 649 & $3.878$       \\
\texttt{DriverQuery} & 1.168 & 68.668 & 1 735 859 & $3.860$            \\
\texttt{PowerCFG} & 61.333 & 273.882 & 4 497 247 & $4.726$       \\
\texttt{7zip}       & 175.526 & 233.792 & 146 578 & $39.751$      \\
\texttt{IZArc}      & 30.920 & 65.770 & 109 142 & $31.931$      \\
\texttt{WinRar} & 155.674 & 215.455 & 84 329 & $70.890$      \\
\texttt{Cipher} & 0.028 & 4.435 & 24 141 & $18.255$           \\
\texttt{OpenSSL} & 24.769 & 1524.763 & 57 410 809 & $2.613$                \\
\texttt{FCIV} & 10.807 & 87.948 & 725 133 &  $10.638$               \\
\texttt{TurboSFV} & 34.412 & 74.657 & 183 455 & $21.937$ 
\\
\hline
\end{tabularx}
\vspace{0.2cm}
\caption{Run-time overhead for benign applications}

@@

Lastly, anti-analysis checks are aimed at detecting the existence of processes belonging to popular analysis tools. One of such tools is Process Monitor (\texttt{procmon.exe}), which is an advanced Windows monitoring tool that shows real-time file system, registry and process/thread activity \cite{procmon}. Al-Khaser looks for the processes of many analysis tools, including the process associated to Process Monitor. Again, BlueTracer was also able to successfully log anti-analysis events, even the ones directed at uncovering the presence of Process Monitor (\textit{Listing 4.11}). From the listing below, it can be seen that the course of action adopted by Al-Khaser is quite straightforward. It first takes a snapshot of all the processes in the system with \texttt{CreateToolhelp32Snapshot}. Then, for each process in the snapshot, it retrieves the corresponding name, through the use of \texttt{Process32FirstW} at the beginning and \texttt{Process32NextW} afterwards, and compares it with \texttt{procmon.exe} via \texttt{StrCmpIW}.
